A Lyapunov Analysis of Momentum Methods \\ in Optimization

### 2016-10-28 LyapDraft.tex
Update LyapDraft.tex

\subsection{Other Discretizations}
We give examples of two other algorithms that are the result of discretizing the momentum dynamic~\eqref{Eq:DynaBeta}, and analyze them using the discretized Lyapunov function~\eqref{Eq:DiscLyapFunc1}. The proofs of these results can be found in the Appendix~\ref{App:OtherDiscret}. The first method is the result of applying explicit Euler method to~\eqref{Eq:X} and the implicit Euler method to~\eqref{Eq:Z}:
\begin{algorithm}[H]
\caption{Method 1}
{\bf Assumptions:} $f$ is smooth and $\X$ is convex and compact. \\
Let $A_0 = 1$, $x_0 = z_0$, $\alpha_k = A_{k+1} - A_k$ and $\tau_k = \frac{A_{k+1}- A_k}{A_{k}}$. Define recursively,
\begin{subequations}\label{Eq:QuasiSub2}
\begin{align}
z_{k+1} &= \arg \min_{z \in \X} \left\{\langle \nabla f(x_k),z \rangle + \frac{1}{\alpha_k} D_h(z, z_k)\right\}\\
x_{k+1} &= \frac{\tau_k}{\tau_k - 1}z_{k+1} + \frac{1}{\tau_k - 1}  x_k
\end{align} 
\end{subequations}
\end{algorithm}
\noindent This algorithm has optimality conditions,
\begin{subequations}
\begin{align}
z_{k+1} &= x_{k+1} - \frac{A_{k}}{A_{k+1} -A_{k}} (x_{k+1} - x_k),\label{Eq:ZSeqOther1}\\
\nabla h(z_{k+1}) -\nabla h(z_k) &= -(A_{k+1} - A_{k})\nabla f(x_{k}) \label{Eq:XSeqOther1}.
\end{align}
\end{subequations}
\noindent Choosing $A_k = k(k+1)/2$ results in an $O(1/k)$ convergence rate. Note, structurally this algorithm is very similar to the conditional gradient method (see~\ref{App:OtherDiscret} for details). The second method does not comport to a straight-forward discretization technique and is given by the updates,
\begin{algorithm}[H]
\caption{Method 2}
{\bf Assumptions:} $f, h$ are convex and differentiable. $h$ is strongly-convex and $f$ has bounded gradients, $\sup_{x\in\X}\|\nabla f(x)\| = G <\infty$. 
Let $A_0 = 1$, $x_0 = z_0$ and $\tau_k = \frac{A_{k+1}- A_k}{A_{k}}$. Define recursively,
\begin{subequations}\label{Eq:QuasiSub3}
\begin{align}
x_{k+1}  &= \frac{\tau_k}{\tau_k + 1} z_k + \frac{1}{\tau_k+1}x_k,\\%\label{Eq:ZSeqQuasi}\\
z_{k+1} &= \arg \min_{z \in \X} \left\{\langle \nabla f(x_{k+1}),z\rangle + \frac{1}{\alpha_k} D_h(z, z_k)\right\}.
\end{align} 
\end{subequations}
\end{algorithm}
\noindent This algorithm has optimality conditions,
\begin{subequations}\label{Eq:QuasiSub3}
\begin{align}
z_{k} &= x_{k+1} + \frac{A_{k}}{A_{k+1} -A_{k}} (x_{k+1} - x_k),\label{Eq:ZSeqMethod2}\\%\label{Eq:ZSeqQuasi}\\
\nabla h(z_{k+1}) -\nabla h(z_k) &= -(A_{k+1} - A_{k})\nabla f(x_{k+1}).\label{Eq:XSeqMethod2}
\end{align} 
\end{subequations}
Notice, \eqref{Eq:QuasiSub3} is very similar to the quasi-monotone subgradient method~\eqref{Eq:QuasiSub1}. In particular, under the same assumptions, it obtains a $O(1/\sqrt{k})$ convergence guarantee. Furthermore, like the quasi-monotone subgradient method, it can be extended to functions which have bounded subgradients instead of full gradients and an additional weighting term can be added.

@@

We briefly point out that the algorithm resulting from applying something like the implicit Euler method to both \eqref{Eq:Frank1}  and \eqref{Eq:Frank2}, 

@@

results in algorithm for which we can show $E_{k+1} - E_k \leq 0$ using~\eqref{Eq:FrankWolfeLyap}. However, like most implicit discretization techniques applied to the momentum dynamics~\eqref{Eq:ELBreg}, this does not lead to an algorithm with manageable subproblems.

@@

\subsection{Discretizing the Dynamic}
As a proof of concept, we begin by analyzing an implicit discretization of the dynamic~\eqref{Eq:StrongDyn}. 

@@

\caption{Implicit-Euler Based Method (Strong Convexity)}

@@

We begin with a brief reivew of the technique of estimate sequences. 
In \cite{Nesterov04}, Nesterov introduced estimate sequences by giving the following definition
\begin{definition}\cite[2.2.1]{Nesterov04} A pair of sequences $\{\phi_k(x)\}_{k=1}^\infty$ and $\{A_k\}_{k=0}^\infty$ $A_k \geq1$ is called an {\em estimate sequence} of function $f(x)$ if 
\[ \frac{1}{A_k} \rightarrow 0 \]
and for any $x \in \R^n$ and all $k \geq 0$, we have 
\begin{equation}\label{Eq:Ineq1}
\phi_k(x) \leq \Big(1 - \frac{1}{A_k}\Big)f(x) + \frac{1}{A_k}\phi_0(x).
\end{equation}
\end{definition}
\noindent The following lemma, given by Nesterov, explains why estimate sequences are useful.
%
%
%Lemma 
\begin{lemma}\label{Lem:Nest2.2.1} \cite[2.2.1]{Nesterov04}
If for some sequence $\{x_k\}_{k\geq0}$ we have 
\begin{equation}\label{Eq:Seq}
f(x_k) \leq \phi_k^\ast \equiv \min_{x \in \X} \phi_k(x),
\end{equation}
then $f(x_k) - f(x^\ast) \leq \frac{1}{A_k} [\phi_0(x^\ast) - f(x^\ast)]$. 
\end{lemma}
%
%
\begin{proof}The proof is simple and can be shown in two lines.
\begin{align*}
f(x_k) \leq \phi_k^\ast \equiv \min_{x \in \X} \phi_k(x) &\overset{\eqref{Eq:Ineq1}}{\leq}\min_{x \in\X}\left[\Big(1 - \frac{1}{A_k}\Big)f(x) + \frac{1}{A_k}\phi_0(x)\right]\\
&\overset{\eqref{Eq:Seq}}{\leq} \Big(1 - \frac{1}{A_k}\Big)f(x^\ast) + \frac{1}{A_k}\phi_0(x^\ast).
\end{align*}
Rearranging gives the desired inequality.
\end{proof}
%
%
Notice, this definition is not at all constructive. Finding sequences which satisfy these conditions is a highly non-trivial task. The next proposition, introduced by Baes in \cite{Baes09} as a slight of extension of Nesterov's Lemma 2.2.2 \cite{Nesterov04}, provides guidance for constructing estimate sequences. Note, this construction is used in \cite{Nesterov04, Nesterov05, Nesterov08, Baes09, Nesterov15, NesterovCond15}, and is the only known way so far to construct an estimate sequence.
\begin{proposition}\cite[2.2]{Baes09}\label{Eq:Prop}
 Let $\phi_0: \X \rightarrow \R$ be a convex function such that $\min_{x \in \X} \phi_0(x)\geq f^\ast$. Suppose also that we have a sequence $\{f_k\}_{k\geq 0}$ of functions form $\X$ to $\R$ that underestimates $f$:
\begin{align}\label{Eq:Under}
f_k(x) \leq f(x) \quad \text{ for all $x \in \X$ and all $k \geq 0$}
\end{align}
Define recursively $A_0 = 1$ 
\begin{align}
\alpha_{k} &:= A_{k+1} - A_k \label{Eq:alpha}\\ 
\tau_k &:= \frac{a_{k}}{A_{k+1}},\label{Eq:tau}
\end{align}
and 
\begin{equation}\label{Eq:Est}
\phi_{k+1}(x) := (1 - \tau_k) \phi_k(x) + \tau_k f_k(x) = \frac{1}{A_{k+1}} \left(A_0\phi_0(x) + \sum_{i=0}^k a_i f_i(x)\right)
\end{equation}
for all $k\geq 0$. Then $\left(\{\phi_k\}_{k\geq0}, \{A_k\}_{k\geq0}\right)$ is an estimate sequence.
\end{proposition}
%\begin{remark}
%Going forth, we will denote the object
%\begin{equation}\label{Eq:Est2}
%\psi_k(x) := \sum_{i=0}^k a_i f_i(x) + \phi_0(x)
%\end{equation}
%as our estimate sequence.

>>>>

\subsection{Other Discretizations}
We give examples of two other algorithms that are the result of discretizing the momentum dynamics~\eqref{Eq:DynaBeta}, and analyze them using the discretized Lyapunov function~\eqref{Eq:DiscLyapFunc1}. The proofs of these results can be found in the Appendix~\ref{App:OtherDiscret}. The first method is the result of applying the explicit Euler method to~\eqref{Eq:X} and the implicit Euler method to~\eqref{Eq:Z}:
\begin{algorithm}[H]
\caption{Method 1}
{\bf Assumptions:} $f$ is smooth and $\X$ is convex and compact. \\
Let $A_0 = 1$, $x_0 = z_0$, $\alpha_k = A_{k+1} - A_k$ and $\tau_k = \frac{A_{k+1}- A_k}{A_{k}}$. Define recursively,
\begin{subequations}\label{Eq:QuasiSub2}
\begin{align}
z_{k+1} &= \arg \min_{z \in \X} \left\{\langle \nabla f(x_k),z \rangle + \frac{1}{\alpha_k} D_h(z, z_k)\right\}\\
x_{k+1} &= \frac{\tau_k}{\tau_k - 1}z_{k+1} + \frac{1}{\tau_k - 1}  x_k
\end{align} 
\end{subequations}
\end{algorithm}
\noindent This algorithm has optimality conditions,
\begin{subequations}
\begin{align}
z_{k+1} &= x_{k+1} - \frac{A_{k}}{A_{k+1} -A_{k}} (x_{k+1} - x_k),\label{Eq:ZSeqOther1}\\
\nabla h(z_{k+1}) -\nabla h(z_k) &= -(A_{k+1} - A_{k})\nabla f(x_{k}) \label{Eq:XSeqOther1}.
\end{align}
\end{subequations}
\noindent Choosing $A_k = k(k+1)/2$ results in an $O(1/k)$ convergence rate. Note that structurally this algorithm is very similar to the conditional gradient method (see~\ref{App:OtherDiscret} for details). The second method does not correspond to a straightforward discretization technique and is given by the updates,
\begin{algorithm}[H]
\caption{Method 2}
{\bf Assumptions:} $f, h$ are convex and differentiable. $h$ is strongly-convex and $f$ has bounded gradients, $\sup_{x\in\X}\|\nabla f(x)\| = G <\infty$. 
Let $A_0 = 1$, $x_0 = z_0$ and $\tau_k = \frac{A_{k+1}- A_k}{A_{k}}$. Define recursively,
\begin{subequations}\label{Eq:QuasiSub3}
\begin{align}
x_{k+1}  &= \frac{\tau_k}{\tau_k + 1} z_k + \frac{1}{\tau_k+1}x_k,\\%\label{Eq:ZSeqQuasi}\\
z_{k+1} &= \arg \min_{z \in \X} \left\{\langle \nabla f(x_{k+1}),z\rangle + \frac{1}{\alpha_k} D_h(z, z_k)\right\}.
\end{align} 
\end{subequations}
\end{algorithm}
\noindent This algorithm has optimality conditions,
\begin{subequations}\label{Eq:QuasiSub3}
\begin{align}
z_{k} &= x_{k+1} + \frac{A_{k}}{A_{k+1} -A_{k}} (x_{k+1} - x_k),\label{Eq:ZSeqMethod2}\\%\label{Eq:ZSeqQuasi}\\
\nabla h(z_{k+1}) -\nabla h(z_k) &= -(A_{k+1} - A_{k})\nabla f(x_{k+1}).\label{Eq:XSeqMethod2}
\end{align} 
\end{subequations}
Notice, \eqref{Eq:QuasiSub3} is very similar to the quasi-monotone subgradient method~\eqref{Eq:QuasiSub1}. In particular, under the same assumptions, it achieves a $O(1/\sqrt{k})$ convergence rate. Furthermore, like the quasi-monotone subgradient method, it can be extended to functions which have bounded subgradients instead of full gradients and an additional weighting term can be added.

@@

We briefly point out that the algorithm resulting from applying an implicit Euler method to both \eqref{Eq:Frank1}  and \eqref{Eq:Frank2}, 

@@

results in an algorithm for which we can show $E_{k+1} - E_k \leq 0$ using~\eqref{Eq:FrankWolfeLyap}. However, like most implicit discretization techniques applied to the momentum dynamics~\eqref{Eq:ELBreg}, this does not lead to an algorithm with tractable subproblems.

@@

\subsection{Discretizing the Dynamics}
As a proof of concept, we begin by analyzing an implicit discretization of the dynamics~\eqref{Eq:StrongDyn}. 

@@

\caption{Implicit Euler Based Method (Strong Convexity)}

@@

We begin with a brief reivew of the technique of estimate sequences. 
In \cite{Nesterov04}, Nesterov introduced estimate sequences by giving the following definition
\begin{definition}\cite[2.2.1]{Nesterov04} A pair of sequences $\{\phi_k(x)\}_{k=1}^\infty$ and $\{A_k\}_{k=0}^\infty$ $A_k \geq1$ is called an {\em estimate sequence} of function $f(x)$ if 
\[ \frac{1}{A_k} \rightarrow 0 \]
and for any $x \in \R^n$ and all $k \geq 0$, we have 
\begin{equation}\label{Eq:Ineq1}
\phi_k(x) \leq \Big(1 - \frac{1}{A_k}\Big)f(x) + \frac{1}{A_k}\phi_0(x).
\end{equation}
\end{definition}
\noindent The following lemma, given by Nesterov, explains why estimate sequences are useful.
%
%
%Lemma 
\begin{lemma}\label{Lem:Nest2.2.1} \cite[2.2.1]{Nesterov04}
If for some sequence $\{x_k\}_{k\geq0}$ we have 
\begin{equation}\label{Eq:Seq}
f(x_k) \leq \phi_k^\ast \equiv \min_{x \in \X} \phi_k(x),
\end{equation}
then $f(x_k) - f(x^\ast) \leq \frac{1}{A_k} [\phi_0(x^\ast) - f(x^\ast)]$. 
\end{lemma}
%
%
\begin{proof}The proof is straightforward:
\begin{align*}
f(x_k) \leq \phi_k^\ast \equiv \min_{x \in \X} \phi_k(x) &\overset{\eqref{Eq:Ineq1}}{\leq}\min_{x \in\X}\left[\Big(1 - \frac{1}{A_k}\Big)f(x) + \frac{1}{A_k}\phi_0(x)\right]\\
&\overset{\eqref{Eq:Seq}}{\leq} \Big(1 - \frac{1}{A_k}\Big)f(x^\ast) + \frac{1}{A_k}\phi_0(x^\ast).
\end{align*}
Rearranging gives the desired inequality.
\end{proof}
%
%
Notice, this definition is not constructive. Finding sequences which satisfy these conditions is generally a non-trivial task. The next proposition, formalized by Baes in \cite{Baes09} as an  extension of Nesterov's Lemma 2.2.2 \cite{Nesterov04}, provides guidance for constructing estimate sequences. This construction is used in \cite{Nesterov04, Nesterov05, Nesterov08, Baes09, Nesterov15, NesterovCond15}, and is the only known way so far to construct an estimate sequence.
\begin{proposition}\cite[2.2]{Baes09}\label{Eq:Prop}
 Let $\phi_0: \X \rightarrow \R$ be a convex function such that $\min_{x \in \X} \phi_0(x)\geq f^\ast$. Suppose also that we have a sequence $\{f_k\}_{k\geq 0}$ of functions form $\X$ to $\R$ that underestimates $f$:
\begin{align}\label{Eq:Under}
f_k(x) \leq f(x) \quad \text{ for all $x \in \X$ and all $k \geq 0$}
\end{align}
Define recursively $A_0 = 1$ 
\begin{align}
\alpha_{k} &:= A_{k+1} - A_k \label{Eq:alpha}\\ 
\tau_k &:= \frac{a_{k}}{A_{k+1}},\label{Eq:tau}
\end{align}
and 
\begin{equation}\label{Eq:Est}
\phi_{k+1}(x) := (1 - \tau_k) \phi_k(x) + \tau_k f_k(x) = \frac{1}{A_{k+1}} \left(A_0\phi_0(x) + \sum_{i=0}^k a_i f_i(x)\right)
\end{equation}
for all $k\geq 0$. Then $\left(\{\phi_k\}_{k\geq0}, \{A_k\}_{k\geq0}\right)$ is an estimate sequence.
\end{proposition}
%\begin{remark}
%Going forth, we will denote the object
%\begin{equation}\label{Eq:Est2}
%\psi_k(x) := \sum_{i=0}^k a_i f_i(x) + \phi_0(x)
%\end{equation}
%as our estimate sequence.
